---
title: "Homework_5_Trashaj"
output: html_document
date: "2023-06-20"
runtime: shiny
editor_options: 
  markdown: 
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Plot $u_t$ against $y_t$ for a DCS-t-location, a Beta-t-GARCH and a
Beta-t-GARCH processes for different values of $\nu = 3, 10, 200$.

In order to visualize better how the response of the score changes for
different extreme observations I'm going to use the shiny library to
make the plot interactive.

a)  For a DCS-t location model we are going to define a function for the
    martingale difference sequence, the function that generates numbers
    from a t-student distribution and a function that simulate the model
    using the function provided in the R-score files uploaded in
    virtuale.

Here the proper martingale difference sequence (MDS) can be defined as

\$ u_t = (1 + \frac{v_t^2}{\nu e^{2\lambda}})\^{-1} v_t \$

where $v_t = y_t - \mu{t|t-1}$

```{r}
martingale_diff_u_t <- function(y, mu_t, varsigma, nu){
  
  u_t <- c((1 / (1 + (y - mu_t)^2/(nu*varsigma)) * (y - mu_t)))
  
  return(u_t)
}
```

Now we are going to define the function for the the random generating
numbers

```{r}
uSTDT_rnd <- function(n, mu, varsigma, nu) {
  
  z <- rt(n, df = nu) 
  y <- numeric()
  for(i in 1:n){
    y[i] <- c(mu + z[i]* sqrt(varsigma) ) 
  }
  
  return(y)
}


```

Here we define the function that simulate the process, the innovations
factor and the updated filters. In particular recall that

$\mu_{t+1 | t} = \omega + \phi\mu_{t|t-1} + ku_t$

```{r}
uDCS_t_model_simulator <- function(T, omega, phi, k, varsigma, nu){
  
  ###Define the Processes
  y   <- array(data = NA, dim = c(T) ) 
  
  ###Define Dynamic Location and Innovations
  mu_t <- array(data = NA, dim = c(T))
  u_t  <- array(data = NA, dim = c(T-1))
  
  ###Initial value for the recursion 
  mu_t[1]   <- omega
  
  ###Generate the first observations of the process
  y[1]   <- uSTDT_rnd(1, mu_t[1], varsigma, nu)
  
  ###Dynamics 
  for (t in 2:T) {
    
    ###Factor Innovations
    u_t[t-1] <- martingale_diff_u_t(y[t-1], mu_t[t-1], varsigma, nu)
    
    ###Updating Filters                    
    mu_t[t]   <- omega + phi * (mu_t[t-1] - omega) + k * u_t[t-1]
    
    ###Generate the observations of the processes
    y[t] <- uSTDT_rnd(1, mu_t[t], varsigma, nu)
  }
  
  ###Make List
  out <- list(y_t_gen          = as.ts(y),
              Dynamic_Location = as.ts(mu_t),
              Innovation_u_t   = as.ts(u_t))
  
  return(out)
}

```

Now we can simulate the process using some initial values

```{r}
T <- 800       # number of observations

# Parameters
omega <- 0     
phi <- 0.8   
k <- 0.8      
varsigma <- 2.5


# Values of nu to iterate over
library(shiny)

ui <- fluidPage(
  sliderInput("nu_slider", "nu", min = 3, max = 200, value = 3), # Slider to select the value of nu
  plotOutput("dcs_plot") # Placeholder for the plot
)

server <- function(input, output) {
  output$dcs_plot <- renderPlot({
    nu <- input$nu_slider # Get the current value of nu from the slider
    simu <- uDCS_t_model_simulator(T, omega, phi, k, varsigma, nu) # Run the simulation with the selected nu value
    
    # Generate the y_t time series
    y_t <- simu$y_t_gen
    
    # Plot the y_t time series with a title indicating the current nu value
    ts.plot(y_t, main = paste("DCS simulation for the time varying location (nu =", nu, ")"))
    
    # Obtain the u_t transformation
    u_t <- simu$Innovation_u_t
    lines(u_t, col = "red")
    
  })
}

shinyApp(ui, server) #it may be required to open the plot in a browser

#Uncomment the lines below to see just the three iterations

#nu_values <- c(3, 10, 200)

#for (nu in nu_values) {
  #simu <- uDCS_t_model_simulator(T, omega, phi, k, varsigma, nu)
  
  # y_t
  #y_t <- simu$y_t_gen
  #dim(y_t)
  #ts.plot(y_t, main = paste("DCS simulation for the time varying location (nu =", nu, ")"))
  
  # u_t
  #u_t <- simu$Innovation_u_t
  #dim(u_t)
  #lines(u_t, col = "red")
#}


```

As we can see as $\nu$ increases, $u_t$ show a more pronounced response
to extreme observations. When $\nu$ is smaller, the extreme observations
has a lower impact on the score.

(b) For a Beta-t-GARCH as we did before we are going to introduce the
    three functions but with different equations.

Here the martingale difference sequence that we are going to use is the
following: $u_t = \frac{(\nu + 1) * y^2}{(\nu-2)*\sigma_t + y^2} -1$

```{r}
mds_scale <- function(y, mu, sigma_t, nu){
  u_t <- c((nu+1)*y^2 / ((nu-2)*sigma_t + y^2) - 1)
  return(u_t)
}
```

Now the data generating process of $y_t$ from a $t$-Student
distribution, recall that for a Beta-t-GARCH process
$y_t = \sigma_{t|t-1} * z_t$

```{r}
uSTDT_rnd_scale_beta <- function(n, sigma_t, nu) {
  z <- rt(n, df = nu) 
  y <- numeric()
  for(i in 1:n){
    y[i] <- c(z[i]* sqrt(sigma_t)) 
  }
  return(y)
}
```

And finally the simulator for the scale. Recall that
$\sigma_{t|t-1}^2 = \omega + \phi * \sigma_{t-1|t-2}^2 + k * \sigma_{t-1|t-2}^2 * u_{t-1}$

```{r}
beta_simulator_scale <- function(T, omega, phi, k, mu, nu){
  
  ###Define the Processes
  y   <- array(data = NA, dim = c(T) ) 
  
  ###Define Dynamic Scale and Innovations
  sigma_t <- array(data = NA, dim = c(T))
  u_t  <- array(data = NA, dim = c(T-1))
  
  ###Initial value for the recursion 
  sigma_t[1]   <- omega
  
  ###Generate the first observations of the process
  y[1]   <- uSTDT_rnd_scale_beta(1, sigma_t[1], nu)
  
  ###Dynamics 
  for (t in 2:T) {
    
    ###Factor Innovations
    u_t[t-1] <- mds_scale(y[t-1], mu, sigma_t[t-1], nu)
    
    ###Updating Filters                    
    sigma_t[t]   <- omega + phi * (sigma_t[t-1]^2) + k * (sigma_t[t-1]^2) * u_t[t-1]
    
    ###Generate the observations of the processes
    y[t] <- uSTDT_rnd_scale_beta(1, sigma_t[t], nu)
  }
  ######################
  ####### OUTPUT #######
  ######################
  
  ###Make List
  out <- list(y_t_gen          = as.ts(y),
              Dynamic_scale = as.ts(sigma_t),
              Innovation_u_t   = as.ts(u_t))
  
  return(out)
}
```

Now we can simulate the process with initial values of the parameters

```{r}
T <- 800       # number of observations
# Parameters
omega <- 1
phi <- 0.5   
k <- 0.2      
mu <- 0  

#uncomment those lines if you want so see the slider 

#import shiny to make interactive visualizations
#library(shiny)

#ui <- fluidPage(
  #sliderInput("nu_slider", "nu", min = 3, max = 200, value = 3), # Slider to select the value of nu
  #plotOutput("beta_t_plot") # Placeholder for the plot
#)

#server <- function(input, output) {
#  output$beta_t_plot <- renderPlot({
#    nu <- input$nu_slider # Get the current value of nu from the slider
#    simu <- beta_simulator_scale(T, omega, phi, k, varsigma, nu) # Run the simulation with the selected nu value
    
    # Generate the y_t time series
#    y_t <- simu$y_t_gen
    
    # Plot the y_t time series with a title indicating the current nu value
#    ts.plot(y_t, main = paste("Beta-t-GARCH simulation for the time varying scale(nu =", nu, ")"))
    
    # Obtain the u_t transformation
 #   u_t <- simu$Innovation_u_t
#    lines(u_t, col = "red")
    
  #})
#}

#shinyApp(ui, server) #it may be required to open the plot in a browser

#comment the lines below to see just the plot with the slider

nu_values <- c(3, 10, 200)

for (nu in nu_values) {
  simu <- uDCS_t_model_simulator(T, omega, phi, k, varsigma, nu)
  
   #y_t
  y_t <- simu$y_t_gen
  dim(y_t)
  ts.plot(y_t, main = paste("DCS simulation for the time varying location (nu =", nu, ")"))
  
   #u_t
  u_t <- simu$Innovation_u_t
  dim(u_t)
  lines(u_t, col = "red")
}


```

Even here we can derive the same conclusion as before.

c)  For the Beta-t-EGARCH process the MDS we are going to use is the
    following:

$u_t = \frac{(\nu +1)y_t^2}{\nu*e^{2*\lambda_{t|t-1} + y_t^2} -1$

```{r}
mds_lambda <- function(y, mu, lambda_t, nu){
  u_t <- c((nu+1)*y^2 / (nu*exp(2*lambda_t) + y^2) - 1)
  return(u_t)
}

```

Recall that in the Beta-t-EGARCH process
$y_t = z_t * e^{\lambda_{t|t-1}}$

```{r}
uSTDT_rnd_lambda <- function(n, lambda_t, nu) {
  z <- rt(n, df = nu) 
  y <- numeric()
  for(i in 1:n){
    y[i] <- c(z[i]* exp(lambda_t)) 
  }
  return(y)
}
```

And finally the simulator for the scale. Recall that
$\lambda_{t|t-1} = \omega + \phi * \lambda_{t-1|t-2} + k * u_{t-1}$

```{r}
simulator_lambda <- function(T, omega, phi, k, mu, nu){
  
  ###Define the Processes
  y   <- array(data = NA, dim = c(T) ) 
  
  ###Define Dynamic Scale and Innovations
  lambda_t <- array(data = NA, dim = c(T))
  u_t  <- array(data = NA, dim = c(T-1))
  
  ###Initial value for the recursion 
  lambda_t[1]   <- (log(omega))/2
  
  ###Generate the first observations of the process
  y[1]   <- uSTDT_rnd_lambda(1, lambda_t[1], nu)
  
  ###Dynamics 
  for (t in 2:T) {
    
    ###Factor Innovations
    u_t[t-1] <- mds_lambda(y[t-1], mu, lambda_t[t-1], nu)
    
    ###Updating Filters                    
    lambda_t[t]   <- omega + phi * lambda_t[t-1] + k * u_t[t-1]
    
    ###Generate the observations of the processes
    y[t] <- uSTDT_rnd_lambda(1, lambda_t[t], nu)
  }
  ######################
  ####### OUTPUT #######
  ######################
  
  ###Make List
  out <- list(y_t_gen          = as.ts(y),
              Dynamic_scale = as.ts(exp(2*lambda_t)),
              Innovation_u_t   = as.ts(u_t))
  
  return(out)
}
```

The simulation with initial number of observations and the values of the
parameters

```{r}
T <- 800       # number of observations
# Parameters
omega <- 1    
phi <- 0.5     
k <- 0.2      
mu <- 0


# Values of nu to iterate over
library(shiny)

ui <- fluidPage(
  sliderInput("nu_slider", "nu", min = 3, max = 200, value = 3), # Slider to select the value of nu
  plotOutput("dcs_plot") # Placeholder for the plot
)

server <- function(input, output) {
  output$dcs_plot <- renderPlot({
    nu <- input$nu_slider # Get the current value of nu from the slider
    simu <- simulator_lambda(T, omega, phi, k, varsigma, nu) # Run the simulation with the selected nu value
    
    # Generate the y_t time series
    y_t <- simu$y_t_gen
    
    # Plot the y_t time series with a title indicating the current nu value
    ts.plot(y_t, main = paste("Beta-EGARCH simulation for the time varying scale (nu =", nu, ")"))
    
    # Obtain the u_t transformation
    u_t <- simu$Innovation_u_t
    lines(u_t, col = "red")
    
  })
}

shinyApp(ui, server) #it may be required to open the plot in a browser

#Uncomment the lines below to see just the three iterations

#nu_values <- c(3, 10, 200)

#for (nu in nu_values) {
  #simu <- uDCS_t_model_simulator(T, omega, phi, k, varsigma, nu)
  
  # y_t
  #y_t <- simu$y_t_gen
  #dim(y_t)
  #ts.plot(y_t, main = paste("DCS simulation for the time varying location (nu =", nu, ")"))
  
  # u_t
  #u_t <- simu$Innovation_u_t
  #dim(u_t)
  #lines(u_t, col = "red")
#}
```

The second request asked to simulate a score driven model for the scale
with parameter specified as I choose. So, let's take a GARCH (1,1) model
with standard normal residuals.

Let's define the proper MDS:

$u_t = \frac{y_t^2}{\sigma_{t|t-1}^2}-1$

```{r}
mds_garch <- function(y, mu, sigma_t){
  
  u_t <- c(((y^2)/(sigma_t)) -1)
  return(u_t)
  
}
```

Now we need to generate the observations, recall that
$y_t = \sigma_{t|t-1}*z_t$

where $z_t \thicksim NID(0,1)$

```{r}
uSTDT_rnd_garch <- function(n, sigma_t) {
  z <- rnorm(n,0,1) 
  y <- numeric()
  for(i in 1:n){
    y[i] <- c(z[i]* sqrt(sigma_t)) 
  }
  return(y)
}
```

Here we implement the function simulator, recall that

$\sigma_{t|t-1}^2 = omega + phi*\sigma_{t|t-1}^2 + k*\sigma_{t|t-1}^2*u_t$

```{r}
simulator_garch <- function(T, omega, phi, k, mu){
  
  ###Define the Processes
  y   <- array(data = NA, dim = c(T) ) 
  
  ###Define Dynamic Scale and Innovations
  sigma_t <- array(data = NA, dim = c(T))
  u_t  <- array(data = NA, dim = c(T-1))
  v <- array(data = NA, dim = c(T-1))
  
  ###Initial value for the recursion 
  sigma_t[1]   <- omega
  
  ###Generate the first observations of the process
  y[1]   <- uSTDT_rnd_garch(1, sigma_t[1])
  
  ###Dynamics 
  for (t in 2:T) {
    
    ###Factor Innovations
    u_t[t-1] <- mds_garch(y[t-1], mu, sigma_t[t-1])
    
    ###Updating Filters                    
    sigma_t[t]   <- omega + phi * sigma_t[t-1] + k * sigma_t[t-1] * u_t[t-1]
    
    ###Generate the observations of the processes
    y[t] <- uSTDT_rnd_garch(1, sigma_t[t])
    
    ##Define the prediction error
    
    v[t-1] <- (y[t-1])^2 - sigma_t[t-1]
  }
  ######################
  ####### OUTPUT #######
  ######################
  
  ###Make List
  out <- list(y_t_gen          = as.ts(y),
              Dynamic_scale = as.ts(sigma_t),
              Innovation_u_t   = as.ts(u_t),
              prediction_error = as.ts(v))
  
  return(out)
}
```

Now we can plot the series

```{r}
T <- 800       # number of observations
omega <- 1    
phi <- 0.5    
k <- 0.5       
mu <- 0        

# Simulation
simulation <- simulator_garch(T, omega, phi, k, mu)

# Plot the series y_t:
y_t <- simulation$y_t_gen  # simulated time series 
ts.plot(y_t, main="Simulated GARCH(1,1)")
```

We can plot the time varying parameter

```{r}
time_v <- simulation$Dynamic_scale
ts.plot(time_v, main="Plot the time varying parameter sigma")

```

We can plot $u_t$ and $v_t$

```{r}
v_t <- simulation$prediction_error
ts.plot(v_t, main="Plot of v_t in black and u_t in red")
u_t <- simulation$Innovation_u_t
lines(u_t, col="red")
```

Now to estimate the models I'm going to use the function defined in the
R-score file uploaded in virtuale.

```{r}
uSTDT_uDCS_t <- function(y, sigma_t, varsigma, nu, log = TRUE){
  
  ulpdf <- (lgamma((nu + 1) / 2) - lgamma(nu / 2) - (1/2) * log(varsigma) -
              (1/2)  * log(pi * nu) - ((nu + 1) / 2) * log(1 + (y - sigma_t)^2 / (nu*varsigma) ))
  
  if(log != TRUE){
    ulpdf <- exp(ulpdf)
  } 
  
  return(ulpdf)
}
```

```{r}
uDCS_t_model_filter <- function(y, theta){
  
  ###Take T
  T <- length(y)
  
  ###Define LogLikelihoods
  dloglik <- array(data = NA, dim = c(T))
  loglik  <- numeric()
  
  ###Parameter Selections Dynamic Scale
  omega <- theta[1]
  phi   <- theta[2]
  k     <- theta[3]
  
  varsigma <- theta[4]
  nu       <- theta[5]
  
  ###Define Dynamic Scale and Innovations
  
  u_t  <- array(data = NA, dim = c(T))
  sigma_t  <- array(data = NA, dim = c(T+1))
  
  ###Initialize Dynamic Scale
  sigma_t[1]   <- (omega)
  
  ###Initialize Likelihood
  dloglik[1] <- uSTDT_uDCS_t(y[1], sigma_t[1], varsigma = varsigma, nu = nu, log = TRUE)
  loglik     <- dloglik[1]
  
  for(t in 2:(T+1)) {
    ###Dynamic Scale Innovations
    u_t[t-1] <- mds_garch(y[t-1], sigma_t[t], varsigma)
    ###Updating Filter                    
    sigma_t[t]   <- omega + phi * sigma_t[t-1] + k * sigma_t[t-1] * u_t[t-1]
    
    if(t < (T+1)){
      ###Updating Likelihoods
      dloglik[t] <- uSTDT_uDCS_t(y[t], sigma_t = sigma_t[t], varsigma = varsigma, nu = nu, log = TRUE)
      loglik     <- loglik + dloglik[t]
    }
  }
  
  ######################
  ####### OUTPUT #######
  ######################
  sigma_t <- ts(sigma_t, start = start(y), frequency = frequency(y))
  u_t  <- ts(u_t, start = start(y), frequency = frequency(y))
  
  ###Make List
  out <- list(Dynamic_Scale= sigma_t,
              Innovation_u_t   = u_t,
              Log_Densities_i  = dloglik,
              Log_Likelihood   = loglik)
  
  return(out)
}
```

```{r}
interprete_uDCS_t_model <- function(dati, param){
  
  ###Take T
  T <- length(dati)
  
  ###Parameter Selections Dynamic Location
  omega <- param[1]
  phi   <- param[2]
  k     <- param[3]
  
  varsigma <- param[4]
  nu       <- param[5]
  
  ###Create a new vector with the parameters
  theta_new <- c(omega, phi, k, varsigma, nu)
  
  #------> Fitness Functions
  fitness <- uDCS_t_model_filter(dati, theta_new)$Log_Likelihood
  
  if(is.na(fitness) | !is.finite(fitness)) fitness <- -1e10
  if(fitness != fitness) fitness <- -1e10
  
  return(-fitness)
}
```

```{r}
garch_estimator <- function(dati, param){
  
  Start <- Sys.time()
  ###Take T
  T <- length(dati)
  
  ###Parameter Selections Dynamic Scale
  omega <- param[1]
  phi   <- param[2]
  k     <- param[3]
  
  varsigma <- param[4]
  nu       <- param[5]
  
  ###Create a vector with the parameters
  theta_st <- c(omega, phi, k, varsigma, nu)
  
  ###Take Bounds
  lower <- c(-Inf, -0.999, -2, 1e-05, 2.099)
  upper <- c( Inf,  0.999,  2, Inf, 300)
  
  #------> Optimize uDCS_t_model Filters w/L-BFGS-B
  #optimizer <- suppressWarnings(optim(par = theta_st, fn = interprete_uDCS_t_model, 
  #                                    dati = dati, method = "L-BFGS-B", 
  #                                    control = list(trace = 1), hessian = FALSE,
  #                                    lower = lower, upper = upper))
  
  #------> Optimize every uDCS_t_model Filters w/solnp := WAY faster than optim
  #optimizer <- suppressWarnings(Rsolnp::solnp(pars = theta_st, fun = interprete_uDCS_t_model, 
  #                                            dati  = dati, control = list(trace = 1), 
  #                                            LB = lower, UB = upper)) 
  
  #------> Optimize every uDCS_t_model Filters w/nlminb := WAY faster than optim
  optimizer <- suppressWarnings(nlminb(start = theta_st, objective = interprete_uDCS_t_model, 
                                       dati  = dati, gradient = NULL, 
                                       control = list(trace = 0), hessian = NULL,
                                       lower = lower, upper = upper))
  
  #------> Save the optimized parameters Dynamic Scale
  omega_opt <- optimizer$par[1]  
  phi_opt <- optimizer$par[2]
  k_opt  <- (optimizer$par[3])
  
  varsigma_opt <- optimizer$par[4]
  nu_opt       <- optimizer$par[5]
  
  ###Create a vector with ALL the optimized parameters
  theta_opt <- c(omega_opt, phi_opt, k_opt, varsigma_opt, nu_opt)
  
  ###Create a list with ALL the optimized parameters
  theta_list <- list(omega = omega_opt,
                     phi = phi_opt,
                     k  = k_opt,
                     varsigma = varsigma_opt,
                     nu    = nu_opt)
  
  ######################
  ####### OUTPUT #######
  ######################
  
  #------> Some detail
  Elapsed_Time <- Sys.time() - Start
  print(paste("Elapsed Time: ", toString(Elapsed_Time)))
  
  ###Make List
  out <- list(theta_list = theta_list,
              theta      = theta_opt,
              optimizer  = optimizer)
  
  return(out) 
}
```

```{r}
# Estimation 
set.seed(1234)
mod_est <- garch_estimator(y_t, c(omega, phi, k, 1, 30))
par<-mod_est$theta_list # estimated parameters 
par
```

```{r}
#Estimation of the sigma
est_sigma<- array(data = NA, dim = T)
est_w<- array(data = NA, dim = T-1)
est_sigma[1] <- 1
for(t in 2:T) {
  est_w [t-1] <- (y_t[t-1]^2 / est_sigma[t-1]^2) - 1
  est_sigma[t]<- par$omega + (par$phi * est_sigma[t-1]^2)+ (par$k * est_sigma[t-1]^2 * est_w[t-1])
}
ts.plot(est_sigma, main="Estimated time varing sigma ", ylim=c(-3,3), xlim=c(0,300))
lines(time_v, col="blue")


```

Here in the last graph we can see that the estimated variance is far
different from the simulated one (blue line). This is because they have
been estimated with a different \$z_t\$ (t-Student and not Normal
distribution).
